Allocation Removal by Partial Evaluation in a Tracing JIT

### 2011-01-19 talk/pepm2011/presentation/talk.tex
tweaks to the slides after the first test run

\author[Bolz et. al.]{\emph{Carl Friedrich Bolz}\inst{1} \and Antonio Cuni\inst{1} \and Maciej Fijałkowski\inst{2} \and Michael Leuschel\inst{1} \and \\
            Samuele Pedroni\inst{3} \and Armin Rigo\inst{1}}
\author{\emph{Carl Friedrich Bolz} \and Antonio Cuni \and Maciej Fijałkowski \and Michael Leuschel \and Samuele Pedroni \and Armin Rigo}
% - Give the names in the same order as the appear in the paper.
% - Use the \inst{?} command only if the authors have different
%   affiliation.

\institute[Heinrich-Heine-Universität Düsseldorf]
{Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany \and

 merlinux GmbH, Hildesheim, Germany \and

 Open End, Göteborg, Sweden \and
}

@@

%\begin{frame}
%  \frametitle{Outline}
%  \tableofcontents
  % You might wish to add the option [pausesections]
%\end{frame}

@@

\begin{frame}
  \frametitle{Dynamic Languages are Slow}
  \begin{itemize}
      \item Interpretation overhead
      \item Constant type dispatching
      \item Boxing of primitive types
      \pause
      \begin{itemize}
          \item A lot of allocation of short-lived objects
      \end{itemize}
  \end{itemize}
\end{frame}

@@

  evaluated in an interpreter:
  \pause
  \begin{enumerate}
      \item What's the type of a? \texttt{Integer}
      \item What's the type of b? \texttt{Integer}
  \pause
      \item unbox a
      \item unbox b
      \item compute the sum
      \item box the result as an \texttt{Integer}
      \item store into x
  \pause
      \item What's the type of x? \texttt{Integer}
      \item What's the type of c? \texttt{Integer}
  \pause
      \item unbox x
      \item unbox c
      \item compute the sum
      \item box the result as an \texttt{Integer}
      \item store into y
  \end{enumerate}
\end{frame}

\begin{frame}
  \frametitle{What to do?}
  \begin{itemize}
      \item Hard to improve in an interpreter
      \item Use a JIT compiler
      \item \textbf{Add a optimization that can deal with heap operations}
      \pause
      \item optimize short-lived objects
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Context: The PyPy Project}
  A general environment for implementing dynamic languages
  \pause
  \begin{block}{Approach}
      \begin{itemize}
          \item write an interpreter for the language in RPython
          \item compilable to an efficient C-based VM
          \pause
          \item (RPython is a restricted subset of Python)
      \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{PyPy's Tracing JIT}
  the feature that makes PyPy interesting:
  \begin{itemize}
      \item a meta-JIT, applicable to many languages
      \item needs a few source-code hints (or user annotations) in the interpreter
      \item JIT is a tracing JIT compiler
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Tracing JITs}
  \begin{itemize}
      \item VM contains both an interpreter and the tracing JIT compiler
      \item JIT works by observing and logging what the interpreter does
      \item for interesting, commonly executed code paths
      \item produces a linear list of operations (trace)
      \item trace is optimized and then turned into machine code
  \end{itemize}
\end{frame}

@@

\begin{frame}
  \frametitle{Tracing JIT: Advantages}
  \begin{itemize}
      \item Traces are interesting linear pieces of code
      \item most of the time correspond to loops
      \item everything called in the trace is inlined
      \item can perform good optimizations on the loop
      \item rarer paths run by the interpreter
  \end{itemize}
\end{frame}

@@

\begin{frame}
  \frametitle{Operations Manipulating Heap Objects}
  \begin{itemize}
      \item \texttt{v = new(T)} makes a new object
      \item \texttt{get(w, F)} reads a field out of an object
      \item \texttt{u = set(v, F, w)} writes a field of an object
      \item \texttt{guard(v, T)} checks the type of an object
  \end{itemize}
\end{frame}

@@

\begin{frame}
  \frametitle{Benchmark Results}
  \begin{itemize}
      \item to evaluate the optimization we used PyPy's Python interpreter with real-world programs
      \item optimization can remove
      \begin{itemize}
          \item 70\% of all \texttt{new} operations
          \item 14\% of all \texttt{get/set} operations
          \item 93\% of all \texttt{guard} operations
      \pause
      \item Timings improve by a factor between 1.2 and 6.95
      \item outperforming standard Python on all benchmarks but one
      \pause
      \item more details in the paper
      \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Conclusion}
  \begin{itemize}
      \item very simple partial-evaluation-based optimization
      \item can remove a lot of allocations and type checks in practical programs
      \item no control issues
      \item all control decisions made by the tracer
      \item based on observing executing program
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Questions?}
  \begin{itemize}
      \item very simple partial-evaluation-based optimization for tracing JITs of dynamic languages
      \item can remove a lot of allocations and type checks in practical programs
      \item no control issues
      \item all control decisions made by the tracing JIT
      \item based on observing executing program
  \end{itemize}
\end{frame}

>>>>

\author[Carl Friedrich Bolz et. al.]{\emph{Carl Friedrich Bolz}\inst{1} \and Antonio Cuni\inst{1} \and Maciej Fijałkowski\inst{2} \and Michael Leuschel\inst{1} \and Samuele Pedroni\inst{3} \and Armin Rigo\inst{1}}
% - Give the names in the same order as the appear in the paper.
% - Use the \inst{?} command only if the authors have different
%   affiliation.

\institute[Heinrich-Heine-Universität Düsseldorf]
{$^1$Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany \and

 $^2$merlinux GmbH, Hildesheim, Germany \and

 $^3$Open End, Göteborg, Sweden \and
}

@@

@@

\begin{frame}
  \frametitle{Dynamic Languages are Slow}
  \begin{itemize}
      \item Interpretation overhead
      \item Type dispatching
      \item Boxing of primitive types
      \pause
      \begin{itemize}
          \item A lot of allocation of short-lived objects
      \end{itemize}
  \end{itemize}
\end{frame}

@@

  evaluated in an interpreter:
  \pause
  \begin{enumerate}
      \item What's the type of a? \texttt{Integer}
      \item What's the type of b? \texttt{Integer}
  \pause
      \item unbox a
      \item unbox b
      \item compute the sum
  \pause
      \item box the result as an \texttt{Integer}
      \item store into x
  \pause
      \item What's the type of x? \texttt{Integer}
      \item What's the type of c? \texttt{Integer}
  \pause
      \item unbox x
      \item unbox c
      \item compute the sum
  \pause
      \item box the result as an \texttt{Integer}
      \item store into y
  \end{enumerate}
\end{frame}

\begin{frame}
  \frametitle{What to do?}
  \begin{itemize}
      \item Use a JIT compiler
      \item \textbf{Add a optimization that can deal with heap operations}
      \pause
      \item optimize short-lived objects
      \item remove some of the redundancy
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Overview}
  \tableofcontents
  % You might wish to add the option [pausesections]
\end{frame}

\section{Experimental Context}

\begin{frame}
  \frametitle{Our Experimental Context is the PyPy Project}
  A general environment for implementing dynamic languages
  \pause
  \begin{block}{Approach}
      \begin{itemize}
          \item write an interpreter for the language in RPython
          \item compilable to an efficient C-based VM
          \pause
          \item (RPython is a restricted subset of Python)
      \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{PyPy Uses a Novel Tracing JIT}
  the feature that makes PyPy interesting:
  \begin{itemize}
      \item a meta-JIT, applicable to many languages
      \item needs a few source-code hints (or user annotations) in the interpreter
      \item JIT is a tracing JIT compiler
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Tracing JITs Compile by Observing an Interpreter}
  \begin{itemize}
      \item VM contains both an interpreter and the tracing JIT compiler
      \item JIT works by observing and logging what the interpreter does
      \item for interesting, commonly executed code paths
      \item produces a linear list of operations (trace)
      \item trace is optimized and then turned into machine code
  \end{itemize}
\end{frame}

@@

\begin{frame}[containsverbatim]
  \frametitle{Optimized Example Trace}
  Trace of \texttt{x = a + b; y = x + c}:
\begin{alltt}
guard_class(a, Integer)
guard_class(b, Integer)
i1 = get(a, intval)
i2 = get(b, intval)
i3 = int_add(i1, i2)
\sout{x = new(Integer)}
\sout{set(x, intval, i3)}
\end{alltt}
\begin{alltt}
\sout{guard_class(x, Integer)}
guard_class(c, Integer)
\sout{i4 = get(x, intval)}
i5 = get(c, intval)
i6 = int_add(\emph{i3}, i5)
y = new(Integer)
set(y, intval, i6)
\end{alltt}
\end{frame}

\begin{frame}
  \frametitle{The Advantages of Tracing JITs}
  \begin{itemize}
      \item Traces are interesting linear pieces of code
      \item most of the time correspond to loops
      \item everything called in the trace is inlined
      \item can perform good optimizations on the loop
      \item rarer paths run by the interpreter
  \end{itemize}
\end{frame}

\section{Proposed Optimization}

@@

\begin{frame}
  \frametitle{Operations Manipulating Heap Objects}
  \begin{itemize}
      \item \texttt{v = new(T)} makes a new object
      \item \texttt{u = get(w, F)} reads a field out of an object
      \item \texttt{set(v, F, w)} writes a field of an object
      \item \texttt{guard(v, T)} checks the type of an object
  \end{itemize}
\end{frame}

@@

\section{Benchmarks}

\begin{frame}
  \frametitle{Benchmark Results}
  \begin{itemize}
      \item to evaluate the optimization we used PyPy's Python interpreter with real-world programs
      \item optimization can remove
      \begin{itemize}
          \item 70\% of all \texttt{new} operations
          \item 14\% of all \texttt{get/set} operations
          \item 93\% of all \texttt{guard} operations
      \pause
      \item Timings improve by a factor between 1.1 and 6.95
      \item outperforming standard Python on all benchmarks but one
      \pause
      \item more details in the paper
      \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Conclusion}
  \begin{itemize}
      \item We propose a very simple partial-evaluation-based optimization for tracing
      JITs of dynamic languages that:
      \begin{itemize}
          \item can remove a lot of allocations and type checks in practical
          programs. 
          \item is efficient and effective.
          \item has no control issues because all control decisions are made by
          the tracing JIT.
      \end{itemize}
      \pause
      \item We claim that this is a general strategy to get rid of control
      problems by simply observing the runtime behaviour of the program.
  \end{itemize}
\end{frame}
